---
title: "Regional Distiction in South Korea"
output: html_document
---
```{r}
# remove all stored data, values, and function
# rm(list = ls())
```

# Basic Info: Source, Library, Data Call
## Source
```{r}
# 인구감소지역 지정: https://www.mois.go.kr/frt/sub/a06/b06/populationDecline/screen.do
# 서울과 세종은 적용 대상 X --> 테스트 포함 안 함
# 2022년 지방소멸대응기금: https://www.korea.kr/news/policyNewsView.do?newsId=148898780
# 지방소멸대응기금 최신: https://www.mois.go.kr/frt/sub/a06/b06/localextinctionFund/screen.do
```
## Library
```{r}
library(httr)
library(httr2)
library(rvest)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyverse)
```

```{r Font}
# Font Load
library(showtext)
library(sysfonts)

font_add_google("Nanum Gothic", "nanum")
showtext_auto()
showtext_opts(dpi = 400) 
#"전출지/전입지(시도)/성/연령(5세)별 이동자수"
```

## Data Call
```{r}
dist_pop_url <- "https://kosis.kr/openapi/Param/statisticsParameterData.do?method=getList&apiKey=NWI0MzQ3OTk3NjRhMTQ1ZjU2ZTUyN2Y4OWRlMjIwNmY=&itmId=T20+T21+T22+&objL1=00+11+26+27+28+29+30+31+36+41+51+43+44+52+46+47+48+50+&objL2=&objL3=&objL4=&objL5=&objL6=&objL7=&objL8=&format=json&jsonVD=Y&prdSe=M&startPrdDe=201806&endPrdDe=202511&orgId=101&tblId=DT_1B040A3"

resp <- GET(dist_pop_url)

dist_pop <- content(resp, as = "text", encoding = "UTF-8") 
dist_pop <- fromJSON(dist_pop)
dist_pop
```

```{r}
dist_entry <- read_csv('5yr_dist_entry_expand.csv',
                       locale = locale(encoding = "EUC-KR")
                       )

dist_entry
```

# Data Cleaning
## dist_pop
### Column erase
```{r}
colnames(dist_pop)
cat('\n')
unique(dist_pop$ORG_ID)
#UNIT_NM_ENG, C1_OBJ_NM, PRD_SE, TBL_ID, TBL_NM, ITM_NM_ENG, C1_NM_ENG, LST_CHN_DE, C1_OBJ_NM_ENG,ORG_ID unnecessary
```

```{r}
dist_pop <- dist_pop |> select(-c(UNIT_NM_ENG, C1_OBJ_NM, PRD_SE, TBL_ID, TBL_NM,
                                  ITM_NM_ENG, C1_NM_ENG, LST_CHN_DE, C1_OBJ_NM_ENG,
                                  ORG_ID))
dist_pop
```

### Date: Year / Month Separation
```{r}
dist_pop$YEAR <- as.integer(substr(dist_pop$PRD_DE, 1, 4))
dist_pop$MONTH <- as.integer(substr(dist_pop$PRD_DE, 5, 6))
dist_pop
```

```{r}
# Zoo library helps the date format into Month and Year, without the whole YMD
library(zoo)
dist_pop$PRD_DE <- as.yearmon(dist_pop$PRD_DE,
                              format = '%Y%m')
dist_pop
```

### Districts Compiled to Physical Regions
```{r}
dist_pop <- dist_pop |>
  mutate(region = case_when(C1_NM == '강원특별자치도'~'강원도',
                            C1_NM %in% c('서울특별시', '경기도', '인천광역시') 
                            ~ '수도권',
                            C1_NM %in% c('충청북도','충청남도','대전광역시','세종특별자치시')
                            ~'충청권',
                            C1_NM %in% c('광주광역시', '전북특별자치도', '전라남도') 
                            ~ '전라권',
                            C1_NM == '제주특별자치도' ~'제주',
                            C1_NM %in% c('경상남도', '경상북도','부산광역시','대구광역시')
                            ~ '경상권',
                            C1_NM == '전국' ~ '전국',
                            TRUE ~ NA_character_)
         )

dist_pop
```

```{r}
unique(dist_pop$ITM_NM)
cat('\n')
unique(dist_pop$UNIT_NM)
# Don't need Men and Woman
# Also units are unnecessary
```

```{r}
dist_pop <- dist_pop |> filter(!ITM_NM %in% c('남자인구수',
                                              '여자인구수')) |>
  select(-c(UNIT_NM))
dist_pop
```

## dist_entry
### Column Cleaning
```{r}
colnames(dist_entry)
```

```{r}
cat("Unique values for 전입지별:\n")
print(unique(dist_entry$전입지별))

cat('\nUnique values for 단위:\n')
unique(dist_entry$단위) #--> units no need
```

```{r}
#월 letter deletion

dist_entry <- dist_entry |> select(-c(`...97`, `단위`)) |> 
  filter(항목 == '순이동자수[명]',
         성별 == '계')
#OR
#dist_entry <- dist_entry %>% select (-...66)
names(dist_entry) <- gsub(" 월$", "", names(dist_entry))

dist_entry
```

### Age group compile
```{r}
dist_entry <- dist_entry |> mutate('연령 계층' = case_when(
  연령별 %in% c("계") ~ '총',
  연령별 %in% c("0 - 4세", "5 - 9세", "10 - 14세") ~ '유소년층',
  연령별 %in% c("15 - 19세", "20 - 24세") ~ '청소년/청년 초기층',
  연령별 %in% c("25 - 29세", "30 - 34세") ~ '청년 후기층',
  연령별 %in% c("35 - 39세", "40 - 44세", "45 - 49세", "50 - 54세") ~ '중년층',
  연령별 %in% c("55 - 59세", "60 - 64세") ~ '장년층',
  연령별 %in% c("65 - 69세", "70 - 74세", "75 - 79세", "80세이상") ~ '고령층',
  TRUE ~ NA_character_
))

dist_entry
```

```{r}
# Age group column relocation
dist_entry <- dist_entry |> relocate('연령 계층', .after = '연령별')
dist_entry
```

```{r}
unique(dist_entry$전입지별)
cat('\n')
unique(dist_pop$C1_NM)
cat('\n')
dist_pop
```

# Visualization
```{r}
dist_pop |> filter(region != '전국') |> 
  group_by(region, YEAR) |>
  summarise(
    avg_pop = mean(as.numeric(DT), na.rm = T),
    .groups = 'drop'
  ) |>
  ggplot(aes(x = YEAR, y = avg_pop, color = region)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = '연도별 평균 인구',
    x = '연도',
    y = '평균 인구 (수)',
    color = '광역권'
  ) + 
  theme_minimal()
```
Regional(not compiled as Metro)
```{r}
#color set by me
my_colors <- c(
  '서울특별시' = '#AE1832',
  '경기도' = '#00983B',
  '인천광역시' = '#0279C2',
  '강원도' = '#9B5069',
  '경상권' = '#35657A',
  '전라권' = '#517A35',
  '제주' = '#8A5C3E',
  '충청권' = '#6D683D'
)

dist_pop |> filter(region != '전국') |>
  mutate(
    region_adj = case_when(
      region == '수도권' & C1_NM %in% c('서울특별시', '경기도', '인천광역시') ~ C1_NM, 
      T ~ region
    )
  ) |>
  group_by(region_adj, YEAR) |>
  summarise(
    avg_pop = mean(as.numeric(DT), na.rm = T),
    .groups = 'drop'
  ) |>
  ggplot(aes(x = YEAR, y = avg_pop, color = region_adj)) +
  geom_point(size = 2) + 
  geom_line(size = 1.2) + 
  scale_color_manual(values = my_colors) + 
  labs(
    title = '연도별 평균 인구',
    x = '연도',
    y = '평균 인구 (수)',
    color = '지역'
  ) + 
  theme_minimal()
```

```{r}
ex1 <- dist_pop |> filter(region != '전국') |>
  mutate(
    region_adj = case_when(
      region == '수도권' & C1_NM %in% c('서울특별시', '경기도', '인천광역시') ~ C1_NM, 
      T ~ region
    )
  ) |>
  group_by(region_adj, YEAR) |>
  summarise(
    avg_pop = mean(as.numeric(DT), na.rm = T),
    .groups = 'drop'
  ) |>
  ggplot(aes(x = YEAR, y = avg_pop, color = region_adj)) +
  geom_point(size = 1.5, alpha = 0.5) + 
  geom_line(size = 1.2) + 
  scale_color_manual(values = my_colors) + 
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ region_adj, scales = "free_y", ncol = 3) + 
  labs(
    title = '연도별 평균 인구',
    x = '연도',
    y = '평균 인구 (수)',
    color = '지역'
  ) + 
  theme(legend.position = 'none', strip.text = element_text(size = 12, face = 'bold')) + 
  theme_minimal()

ex1

#ggsave("Example plot1.png", plot = ex1, width = 14, height = 10, dpi = 300)
```

```{r}
pop_df <- dist_pop |>
  group_by(region, YEAR) |>
  summarise(
    avg_pop = mean(as.numeric(DT), na.rm = TRUE),
    .groups = "drop"
  )

plot_fun <- function(df, title, ncol){
  ggplot(df, aes(x = YEAR, y = avg_pop)) +
    geom_line(linewidth = 1.3) +
    geom_point(size = 2, alpha = 0.7) +
    facet_wrap(~ region, scales = "free_y", ncol = ncol) +
    scale_x_continuous(breaks = 2018:2025) +
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
    labs(title = title, x = "Year", y = "Avg Population (M)") +
    theme_minimal()
}


region_dfs <- pop_df |> filter(region != '전국')

p3 <- plot_fun(region_dfs, "Avg Pop per Year (Metro & Non-Metro)", ncol = 2)

p3

# ggsave("Ex_Plot_Metro and NonMetro.png", width = 12, height = 8, dpi = 300)
```

# Geographic Location Visualization
```{r}
library(sf)
shp <- st_read('ctprvn_20230729/ctprvn.shp', options = "ENCODING=CP949")
shp
```

```{r}
str(shp)
```
# Sample ggplot
```{r}
ggplot(shp) +
  geom_sf(fill = 'grey90', color = 'grey30') + #--> 괄호 안에는 뒷 그리드 없애는 용도
  theme_void()
```
# Data Rearrange for net migration rate standardization
```{r}
dist_entry_edit <- dist_entry |>
  pivot_longer(cols = matches("^\\d{4}.\\d{2}$"),
               names_to = 'DATE',
               values_to = '순이동자수[명]'
               ) |>
  mutate(
    YEAR = as.integer(str_sub(DATE, 1, 4)),
    MONTH = as.integer(str_sub(DATE, 6, 7))
  ) |>
  select(-DATE)

dist_entry_edit
```

```{r}
# erase 전국
dist_entry_edit <- dist_entry_edit |> filter(전입지별 != '전국')
dist_entry_edit
```

```{r}
population_clean <- dist_pop |>
  mutate(population = as.numeric(DT)) 
  # |>select(region, YEAR, MONTH, population)

population_clean
```

```{r}
#now scale the net migration
dist_entry_scaled <- dist_entry_edit |>
  rename(`C1_NM` = `전입지별`) |>
  left_join(population_clean, by = c("C1_NM", "YEAR", "MONTH")) |>
  mutate(
    migration_rate = `순이동자수[명]` / `population`,
    migration_per_1000 = migration_rate * 1000
  )
dist_entry_scaled
```

```{r}
unique(dist_entry_scaled$전출지별)
dist_entry_scaled <- dist_entry_scaled |> select(-c(전출지별, ITM_ID))

dist_entry_scaled
```

```{r}
#find if there's mismatched row
dist_entry_scaled |> summarise(
  n = n(),
  missing_pop = sum(is.na(population))
)
```

# Color the net migration rate
```{r}
# delta_city_means 사용
# viz_dist <- dist_entry_scaled |>
#   group_by()
#dist_entry_scaled

viz_dist <- dist_entry_scaled |> 
  mutate(period = ifelse(
    (YEAR < 2022) | (YEAR == 2022 & MONTH == 1),
    'Pre', 'Post'
  )) |>
  group_by(C1_NM, period) |> 
  summarise(mean_mig_1000 = mean(migration_per_1000), .groups = 'drop')

viz_dist
```

```{r}
viz_dist <- viz_dist |> pivot_wider(
    names_from = period,
    values_from = mean_mig_1000
  )

viz_dist
```

```{r}
names(shp)
unique(shp$CTP_KOR_NM)
```

```{r}
#Jeonbuk name change reflect
shp$CTP_KOR_NM[shp$CTP_KOR_NM == '전라북도'] <- '전북특별자치도'
shp
```

```{r}
shp_join <- shp |>
  left_join(viz_dist, by = c('CTP_KOR_NM' = 'C1_NM'))

shp_join
```

```{r}
lim <- range(c(shp_join$Pre, shp_join$Post), na.rm = TRUE)

equi_scale <- scale_fill_gradient2(
  low = '#d73027',
  mid = 'grey',
  high = 'blue',
  midpoint = 0,
  limits = lim,
  name = 'Net Migration\n(per thousands)',
  guide = guide_colorbar(
    title.position = 'top',
    title.hjust= 0.5
  )
)

# Pre
pre_geo <- ggplot(shp_join) +
  geom_sf(aes(fill = Pre), color = 'white', size = 0.2) + 
  equi_scale + 
  labs(title = 'Before Policy') + 
  theme_void() 

pre_geo
# ggsave("Pre Geo.png", plot = pre_geo
#        , width = 12, height = 10, dpi = 300
#        )

#Post
post_geo <- ggplot(shp_join) +
  geom_sf(aes(fill = Post), color = 'white', size = 0.2) + 
    equi_scale +
  labs(title = 'After Policy') + 
  theme_void()

post_geo
# ggsave("Post Geo.png", plot = post_geo 
#        , width = 12, height = 10, dpi = 300
#        )
```
# t-test & ANVOA
## t-test
```{r}
city_means <- dist_entry_scaled |>
  filter(!(C1_NM %in% c('서울특별시', '세종특별자치시'))) |>
  mutate(period = ifelse(
    (YEAR < 2022) | (YEAR == 2022 & MONTH == 1),
    'Pre', 'Post'
  )) |>
  group_by(C1_NM, period) |>
  summarise(mean_mig_1000 = mean(migration_per_1000), .groups = 'drop')

city_means
```

```{r}
t.test(
  city_means$mean_mig_1000[city_means$period == 'Post'],
  city_means$mean_mig_1000[city_means$period == 'Pre'],
  paired = TRUE
)
```

## ANOVA
```{r}
delta_city_means <- city_means |>
  pivot_wider(
    names_from = period,
    values_from = mean_mig_1000
  ) |>
  mutate(diff = Post - Pre)

delta_city_means
```

# Here, compare regional types (광역시, 도, 특별자치도)
```{r Compare between regional types (광역시, 도, 특별자치도)}
delta_city_means <- delta_city_means |>
  mutate(
    region_type = case_when(
      C1_NM %in% c('강원특별자치도', '전북특별자치도', '제주특별자치도') ~ '특별자치도',
      C1_NM %in% c('경기도', '경상남도', '경상북도', '전라남도', '충청남도', '충청북도') ~ '도',
      TRUE ~ '광역시'
    )
  )

delta_city_means
```

```{r}
#ANOVA 테스팅
anova_res <- aov(diff ~ region_type, data = delta_city_means)
summary(anova_res)
```